Oak resilience to drought and land-use show how the ecological and geographical rear-edges do not necessarily meet today

### 2019-11-05 man/ecosystems/ms_ecosystems.Rmd
update text

## Materials and methods 
### Tree species and study site 
*Melojo* oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate. *Q. pyrenaica* is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation. It is a submediterranean species, hence with lower drought tolerance than genuine evergreen Mediterranean taxa [@delRio2007].

@@

### Greenness data to assess ecosystem resilience
Vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS sensor. EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. MODIS EVI Data were obtained for the period 2000 - 2016. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].

Annual mean EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016. The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about the magnitude of anomalies [@Samanta2012]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$, 
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samanta2012].

@@

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought induced neighbor mortality) in the past. Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median.
```{r, eval=FALSE, echo=FALSE}
# Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004].
```
Site disturbance chronologies were constructed by averaging the individual disturbances series annually. To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014]. In addition, the forest and management history of our sampling sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of existing abundant historical documents was done to compile information about socio-economical activities affecting the studied forests (Table S4).

@@

### Assessing ecosystem and tree individual resilience to drought
To evaluate the effects of drought events on ecosystem resilience (using greenness data) and tree individual resilience (using BAI data) we used resilience indices proposed by @Lloret2011. The Resistance index estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred. The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity. Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels. The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3b) and this length was used in other studies [*e.g.* @Gazol2018]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Figure 5, Table S3).

@@

### Analysis of radial growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 3) and despite competition measured as plot basal area was greatest in CA-High (Table 1). Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 3 and S3a). Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 3), being much more pronounced for the high-elevation, hence more humid, site (CA-High).

Drought events reduced radial growth for all sites (Figure S2a), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies. Tree-growth reduction followed a moisture-gradient. The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2a), being the weakest for the wettest site (Figure 4).

The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940s (the most evident) and the second in 1995-2000. These periods alternated with periods of suppression. In contrast, the two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.

### Resilience to drought events at the ecosystem and individual tree levels
*Rs* and *Rt* varied in the same direction whereas *Rc* varied inversely to *Rs* and *Rt*. Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and *Rc*, significant for all oak populations (Figure 5). A similar pattern was obtained for *Rs* but only significant for SJ site (Figure 5). Importantly, non-significant patterns were obtained if we excluded 1995, except for *Rc* in SJ (Figure S6). The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), particularly SJ where our results suggest a major release event also after 1995 (Figure 4).

During the last two drought events, resilience metrics for greenness and tree-growth were significantly different between drought events (Table 3). The 2005 drought event reduced greenness and growth more than that of 2012 (Tables S1 and S2) but the metrics of resilience generally covaried in the same direction those two years. We obtained significantly higher Resilience and Resistance values for 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 6). Thus, recovery values for greenness were higher for 2005 than for the 2012 drought event (Table S1, Figure 6). Recovery showed a contrasting pattern for EVI and tree-growth.

The resilience metrics calculated significantly varied between sites, except for Resilience of tree-growth (p = 0.534; Table 3), which was similar among the three sites. The two southern populations showed lower values for Recovery than the northern site both for greenness and tree-growth (Tables S1 and S2) but Resistance and Resilience values were significantly higher for the southern site (Tables S1 and S2).

### Tree-growth response to climate
The response of tree-growth to water availability was greater than that to temperatures, as it can be expected for a Mediterranean site and particularly a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure 7a). Yet there were sone differences between northern and southern populations: the positive relationship with SPEI was highest in the northern population (r > 0.6 *vs.* r < 0.5; Figure 7a). In addition, spring maximum temperature was the most significant limitation for tree-growth only for the southern populations (Figure 7b and 7c), whereas minimum and maximum temperatures of current September positively influenced tree-growth only in the northern population.

## Discussion
### Land-use legacies shape sensitivity to climate change of forests and the present rear-edge
The Mediterranean landscapes have suffered a strong transformation driven by human activities throughout history even at high elevations [@GarciaAlix2017]. The review of historical documents shows how forest clearings, firewood extraction, charcoal exploitations and mining have strongly impacted the forests at Sierra Nevada (Table S4) where in addition it has been estimated a historical loss of about 90% of broadleaf *Quercus* species cover from medium and low elevations [@JimenezOlivencia2015]. Therefore, it seems logic to think that the geographical limit of the current rear-edge and relict stands of the deciduous oak studied does not necessarily match the potential dsitribution limit related to edaphoclimatic constraints (*i.e.* the ecological rear-edge). Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland). It has been described a sequential distribution of historical land uses along the elevational gradient in the northern slopes of Sierra Nevada (*e.g.* SJ site): grasslands and shrublands for cattle farming were located at the highest elevations; then forest stands with some croplands; and at the lowest elevations, irrigated terraces with tree crops [@JimenezOlivencia2015]. In addition, other activities like mining should have modified the forest structure at that northern location where there is with a high concentration of mines and quarries that were exploited intermittently throughout history. On the other hand, woodlands in the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks occur [@JimenezOlivencia2015]. Firewood, charcoal extraction and acorns have been resources intensively exploited in southern sites, until at least the middle of the 20th century when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013].

Some of the previous historical records of land-use were likely expressed by the growth chronologies. The release event in 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990]. This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth of the remnant trees included in the northern site (Figure S2b). The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S2b). There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J. Cano-Manuel *personal communications*). Therefore this release might be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004]. On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014]. For southern sites, the only positive release event showed at the earliest years for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].  In this site we observed an increase in growth especially in high elevation trees, where moisture is less limiting, a pattern which concurs with a forest expansion into marginal abandonded croplads and an increase in tree-cover [@CamachoOlmedo2002].

@@

### Relict oaks show high resilience to recent drought events and long-term climatic variability 
Severe drought negatively affects both primary and secondary growth of *Q. pyrenaica* forests. This was expressed by the observed reduction in greenness and tree-growth in response to the 2005 and 2012 drought events, and the consistent radial growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic precipitation negative anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018]. Although the south of Iberian Peninsula suffered severe drought events in the last decades (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto]. For tree-growth, we also observed positive trends in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along its distribution range only at high elevation sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014]. Importantly, for rear-edges where climate change is a threat we would have expected negative growth trends as shown for temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017]. Therefore there is much small-scale variability in the response to climate within the rear-edge we had "a priori" considered in our study, where we even observed positive effects of climate change in certain stands. This suggests that the rear-edge needs to be redefined in consequence [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear-edge) and the potential ecological (limiting) rear-edge of species.

Despite that 2012 drought event was more severe and intense than that in 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012. This could be explained by the different timing of the two droughts. The 2012 drought was a winter drought [@Trigo2013] occurring earlier than the 2005 drought and although the latter lasted less than the 2012 drought, it matched the period of maximum growth for *Q. pyrenaica* forests in late spring (Figure S7). Thus, as bserved for other species [*e.g.* Holm oak; @Camarero2015b] this highlights the importance of the timing of drought as a key factor determining the tree recovery after drought [@Huang2018]. For tree-growth, the highest values of resilience were obtained for the two most severe events, namely 1995 and 1999 (Table S3) and we observed a positive relationship between the tree-growth resilience and drought severity (Figures 6). This result seems to contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events. Yet, the high resilience values obtained for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J. Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of tree resilience estimated from individual growth series.

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of species [@Hampe2005]. A high vulnerability to drought is often assumed for populations located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts. Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability showed by rear-edge populations. Our results would highlight the ample small-scale variability at the ecological boundary and need to better confine the rear-edge limit in our studied forest. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013; @Valbuena2017] seem to be indicative of high local adaptation of this oak. In addition the local environmental conditions of the sites where this species inhabits at Sierra Nevada would explain the low vulnerability to drought showed by this oak. It needs to be mentioned that we only studied adult individuals established decades or centuries ago, meaning that it needs to be assessed if the species expresses resilience or to the contrary vulnerability if we analyzed seedling regeneration, likewise many other Mediterranean species at their xeric limit (2-3 CITAS). A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018; @VilaCabrera2019]. In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q. pyrenaica* in Sierra Nevada are located in a geographical but not a climatic rear-edge. And the current niche is a result of land-use changes, which further complicates the definition of potential rear-edges. Similarly, high values of resilience showed by other Mediterranean species located in their rear-edge [*e.g.* Pinus sylvestris, @Herrero2014] seem to be related with stabilizing mechanisms promoting community resilience that can buffer the impacts of extreme events.

@@

Greenness and tree-growth were more affected by drought events in drier northern populations than in more humid southern oak populations, *e.g.* the northern site showed higher browning intensity than the southern sites during the 2005 drought event and the stronger correlations of tree-growth with SPEI (Hydrological and summer) in the northern site can be interpreted as higher sensitivity to drought in drier sites [@GeaIzquierdo2014]. Similar to many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in the Iberian Peninsula [@GeaIzquierdo2014]. Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @GeaIzquierdo2011; @GarciaGonzalez2017].

>>>>

## Materials and methods 
### Tree species and study site 
*Quercus pyrenaica* Willd. (*melojo* oak) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate. *Q. pyrenaica* is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation. It is a submediterranean species, hence with lower drought tolerance than genuine evergreen Mediterranean taxa [@delRio2007].

@@

### Greenness data to assess ecosystem resilience
Vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS sensor. EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. MODIS EVI data were obtained for the period 2000 - 2016. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].

Annual mean EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016. The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about the magnitude of anomalies [@Samanta2012]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= (\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}})\Big/\sigma_{\mathrm{ref}}$, 
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samanta2012].

@@

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought induced neighbor mortality) in the past. Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median.
```{r, eval=FALSE, echo=FALSE}
# Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004].
```
Site disturbance chronologies were constructed by averaging the individual disturbances series annually. To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014]. In addition, the forest and management history of our sampling sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of existing abundant historical documents was done to compile information about socio-economical activities affecting the studied forests (Table S4).

@@

### Assessing ecosystem and tree individual resilience to drought
To evaluate the effects of drought events on ecosystem resilience (using greenness data) and tree individual resilience (using BAI data) we used resilience indices proposed by @Lloret2011. The Resistance index estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred. The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity. Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels. The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3b) and this length was used in other studies [*e.g.* @Gazol2018]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Table S3).

@@

### Analysis of radial growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites and despite competition measured as plot basal area was greatest in CA-High (Table 1, Figure 3). Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 3 and S3a). Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 3), being much more pronounced for the high-elevation, hence more humid, site (CA-High).

Drought events reduced radial growth for all sites (Figure S2a), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies. Tree-growth reduction followed a moisture-gradient. The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2a), being the weakest for the wettest site (Figure 3).

The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940s (the most evident) and the second in 1995-2000 (Figure 4). These periods alternated with periods of suppression. In contrast, the two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.

### Resilience to drought events at the ecosystem and individual tree levels
*Rs* and *Rt* varied in the same direction whereas *Rc* varied inversely to *Rs* and *Rt*. Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and *Rc*, significant for all oak populations (Figure 5). A similar pattern was obtained for *Rs* but only significant for SJ site. Importantly, non-significant patterns were obtained if we excluded 1995, except for *Rc* in SJ (Figure S5). The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area, particularly SJ where our results suggest a major release event also after 1995 (Figure 4).

During the last two drought events, resilience metrics for greenness and tree-growth were significantly different between drought events (Table 3). The 2005 drought event reduced greenness and growth more than that of 2012 (Tables S1-S2) but the metrics of resilience generally covaried in the same direction those two years. We obtained significantly higher resilience and resistance values for 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 6). Thus, recovery values for greenness were higher for 2005 than for the 2012 drought event (Table S1, Figure 6). Recovery showed a contrasting pattern for EVI and tree-growth.

The resilience metrics calculated significantly varied between sites, except for resilience of tree-growth (p = 0.534; Table 3), which was similar among the three sites. The two southern populations showed lower values for recovery than the northern site both for greenness and tree-growth but resistance and resilience values were significantly higher for the southern site (Tables S1-S2).

### Tree-growth response to climate
The response of tree-growth to water availability was greater than that to temperatures, as it can be expected for a Mediterranean site and particularly a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure 7a). Yet there were some differences between northern and southern populations: the positive relationship with SPEI was highest in the northern population (r > 0.6 *vs.* r < 0.5; Figure 7a). In addition, spring maximum temperature was the most significant limitation for tree-growth only for the southern populations (Figure 7b and 7c), whereas minimum and maximum temperatures of current September positively influenced tree-growth only in the northern population.

## Discussion
### Land-use legacies shape sensitivity to climate change of forests and the present rear-edge
The Mediterranean landscapes have suffered a strong transformation driven by human activities throughout history even at high elevations [@GarciaAlix2017]. The review of historical documents shows how forest clearings, firewood extraction, charcoal exploitations and mining have strongly impacted the forests at Sierra Nevada (Table S4) where in addition it has been estimated a historical loss of about 90% of broadleaf *Quercus* species cover from medium and low elevations [@JimenezOlivencia2015]. Therefore, it seems logic to think that the geographical limit of the current rear-edge and relict stands of the deciduous oak studied does not necessarily match the potential distribution limit related to edaphoclimatic constraints (*i.e.* the ecological rear-edge). Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland). It has been described a sequential distribution of historical land uses along the elevational gradient in the northern slopes of Sierra Nevada (*e.g.* SJ site): grasslands and shrublands for cattle farming were located at the highest elevations; then forest stands with some croplands; and at the lowest elevations, irrigated terraces with tree crops [@JimenezOlivencia2015]. In addition, other activities like mining should have modified the forest structure at that northern location where there is with a high concentration of mines and quarries that were exploited intermittently throughout history. On the other hand, woodlands in the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks occur [@JimenezOlivencia2015]. Firewood, charcoal extraction and acorns have been resources intensively exploited in southern sites, until at least the middle of the 20th century when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013].

Some of the previous historical records of land-use were likely expressed by the growth chronologies. The release event in 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral. This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth of the remnant trees included in the northern site (Figure S2b). The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 4, S2b). There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J. Cano-Manuel *personal communications*). Therefore this release might be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004]. On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014]. For southern sites, the only positive release event showed at the earliest years for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].  In this site we observed an increase in growth especially in high elevation trees, where moisture is less limiting, a pattern which concurs with a forest expansion into marginal abandonded croplads and an increase in tree-cover [@CamachoOlmedo2002].

@@

### Relict oaks show high resilience to recent drought events and long-term climatic variability 
Severe drought negatively affects both primary and secondary growth of *Q. pyrenaica* forests. This was expressed by the observed reduction in greenness and tree-growth in response to the 2005 and 2012 drought events, and the consistent radial growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic precipitation negative anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018]. Despite the severe drought events suffered in the last decades (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto]. For tree-growth, we also observed positive trends in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along its distribution range only at high elevation sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014]. Importantly, for rear-edges where climate change is a threat we would have expected negative growth trends as shown for temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017]. Therefore there is much small-scale variability in the response to climate within the rear-edge we had "a priori" considered in our study, where we even observed positive effects of climate change in certain stands. This suggests that the rear-edge needs to be redefined in consequence [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear-edge) and the potential ecological (limiting) rear-edge of species.

Despite that 2012 drought event was more severe and intense than that in 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012. This could be explained by the different timing of the two droughts. The 2012 drought was a winter drought [@Trigo2013] occurring earlier than the shorter 2005 drought. The latter matched the period of maximum growth for oak forests in late spring (Figure S4). This highlights the importance of the timing of drought as a key factor determining the tree recovery after drought [@Camarero2015b; @Huang2018]. For tree-growth, the highest values of resilience were obtained for the two most severe events (1995 and 1999)(Table S3) and we observed a positive relationship between the tree-growth resilience and drought severity (Figure 5). This result seems to contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events. Yet, the high resilience values obtained for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J. Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of tree resilience estimated from individual growth series.

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of species [@Hampe2005]. A high vulnerability to drought is often assumed for populations located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts. Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability showed by rear-edge populations. Our results would highlight the ample small-scale variability at the ecological boundary and need to better confine the rear-edge limit in our studied forest. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013; @Valbuena2017] seem to be indicative of high local adaptation of this oak. In addition the local environmental conditions of the sites where this species inhabits at Sierra Nevada would explain the low vulnerability to drought showed by this oak. All this, together with the characteristic resprouting ability of the species from its stoloniferous root system, would suggest a long-term persistence of those populations. It needs to be mentioned that we only studied adult individuals established decades or centuries ago, meaning that it needs to be assessed if the species expresses resilience or to the contrary vulnerability if we analyzed seedling regeneration, likewise many other Mediterranean species at their xeric limit (2-3 CITAS). A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018; @VilaCabrera2019]. In this sense, the high values of resilience observed, could suggest that the populations of *Q. pyrenaica* in Sierra Nevada are located in a geographical but not a climatic rear-edge. And the current niche is a result of land-use changes, which further complicates the definition of potential rear-edges. Similarly, high values of resilience showed by other Mediterranean species located in their rear-edge [*e.g.* *Pinus sylvestris*, @Herrero2014] seem to be related with stabilizing mechanisms promoting community resilience that can buffer the impacts of extreme events.

@@

Greenness and tree-growth were more affected by drought events in drier northern populations than in more humid southern oak populations, *e.g.* the northern site showed higher browning intensity than the southern sites during the 2005 drought event and the stronger correlations of tree-growth with SPEI (Hydrological and summer) in the northern site can be interpreted as higher sensitivity to drought in drier sites [@GeaIzquierdo2014]. Similar to many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in the Iberian Peninsula [@GeaIzquierdo2014]. Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017].
